Survey of Cache Oblivious B-trees

### 2011-12-13 paper.tex
Edits

\usepackage{paralist}
\usepackage{graphicx}
\usepackage{url}

@@

\caption{An illustration of the van Emde Boas (vEB) layout}
\label{fig:veb}
\end{figure}

@@

The definition of a SWBST given in \cite{swbst} provides this bound for the
weight of a node $u$:
\begin{equation}
  \frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}
\end{equation}

@@

First, the PMA needs to maintain open space in its memory for newly inserted
items. The PMA is able to handle an insertion into the SWBST using only $O(1 +
(\log^2{N}) / B)$ memory transfers (this is the cost of updating the memory
layout of the SWBST in the PMA, not the cost of updating the data structure
itself). Doing so may seem contradictory: maintaining the locality of elements
while keeping space available for new inserts. The PMA uses a technique
described in \cite{packedmemoryarray}. At a high level, the PMA maintains
windows into the array. When a window becomes too full or too empty, the
elements are spread out within the context of a larger window. Two parameters
are required: the window size and the threshold for triggering a spread out.
However, these parameters are not related to the block size of the memory level
and thus do not violate the CO model.

Second, the PMA needs to handle splits and merges caused by inserting into the
SWBST discussed in the previous section. The PMA is able to accomplish this
using $O(1 + (\log{N}) /B)$ memory transfers. The proof is based on the fact
that each node has $\Theta(c^h)$ descendants (recall, this is the key property
that SWBST provided). Thus, merging or splitting a node will cause
$\Theta(c^h)$ nodes to move, requiring $O(1 + c^h/B)$ memory transfers. It
follows from the definition of SWBST in \cite{swbst} that when a node is
rebalanced, it will require $\Theta(c^h)$ insertions or deletions until it is
rebalanced again. Thus, the amortized cost of each insertion and/or deletion is
$O(1/B)$. When a node is split, it may traverse all the way up the tree, and
thus the split/merge bound is $O((\log N) / B)$ memory transfers.

@@

In \cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
Oblivious String B-tree (COSB-tree), which they argue works more efficiently
for variable length string keys. They also provide a modified front-compression
scheme, which lets their data structure use less space and have improved
compressed data locality.

@@

The dynamic COSB-tree consists of three major components: a centroid tree at
the top, a PMA structure of hash data for the keys known as the \emph{hashdata}
layer; and on the bottom a PMA structure for the original key values,
compressed using an augmented version of forward compression known as the
\emph{keydata} layer. Each leaf of the centroid tree contains a pointer into
the hashdata array. The centroid tree only contains a subset of the keys, so
each leaf points to the respective \emph{neighborhood} of the key being
searched for. After traversing the centroid tree and following the pointer into
the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
sequential scan is performed in the hashdata array layer to find $k$. Each
entry in the hashdata array contains the hash value of the corresponding key,
and a pointer to the compressed key value in the keydata layer.

@@

\subsubsection{The Dictionary Matching Problem and Centroid Trees} %todo:cite dictionary matching problem--sources in cosb-tree paper
\begin{figure}

@@

The centroid tree in the COSB-tree is layed out in memory using a modified van
Emde Boas (vEB) layout for weight-balanced binary trees. The
\emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into
equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x is
denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to
the nearest power of a power of 2. This ensures trees of equal size on all
\emph{levels of detail}, as the centroid tree is traversed.
% Should explain what "level of detail" means here

@@

@@

\subsubsection{Modifying Front Compression to Preserve Locality}

@@

\subsection{Operators}

@@

While B-trees achieve an optimal bound for the number of memory transfers for
search, they do not perform as well for bulk updates and insertions. The
buffered-repository tree (BRT) is an alternative to the B-tree that achieves an
improved amortized $O(\log{\frac{N}{B}})$ memory transfers for inserts compared
to the $O(\log_{B}{N})$ memory access of B-trees. In practice, log-structured
merge (LSM) trees \cite{ONeil96} are used for high-insert workloads.

@@

\subsubsection{Operators}

@@

\Search{} uses $O(\log_{B}{N})$ memory transfers where $N = \Theta(c^{F_k})$ and
$c$ is the fanout. This bound is achieved by searching over the
top and bottom recursive subtrees at each level when traversing the tree. A
key trick is that when you visit a node you only search its largest buffer
(the last buffer in the linked list for that level). The smaller buffers will
be visited eventually as you recurse down the tree. Eventually you will get
down to a single subtree that can fit into a single blocks and thus can be
searched with one memory accesses. Since the fanout is proportional to the
block size it will take us $O(\log_B N)$ block accesses until we get down to a
subtree and thus, by induction, the worst case runtime is $O(\log_B N)$.

>>>>

\usepackage{paralist}
\usepackage{graphicx}
\usepackage{url}
\usepackage{amsmath}

@@

\caption{An illustration of the recursive van Emde Boas (vEB) layout}
\label{fig:veb}
\end{figure}

@@

The definition of a SWBST given in \cite{swbst} provides this bound for the
weight of a node $u$:
\begin{equation}
  \label{eq:bounds}
  \frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}
\end{equation}

@@

First, the PMA needs to maintain open space in its memory for newly inserted
items. Doing so may seem contradictory: maintaining the locality of elements
while keeping space available for new inserts. The achieve a balance between
these extremes the PMA uses a technique
described in \cite{packedmemoryarray}.
The PMA is able to handle an insertion into the SWBST using only $O(1 +
(\log^2{N}) / B)$ memory transfers (this is the cost of updating the memory
layout of the SWBST in the PMA, not the cost of updating the data structure
itself). At a high level, the PMA maintains
``windows'' into the array. When a window becomes too full or too empty, the
elements are spread out within the context of a larger window. Two parameters
are required: the window size and the threshold for triggering a spread out.
However, these parameters are not related to the block size of the memory level
and thus do not violate the CO model.

Second, the PMA needs to handle splits and merges caused by inserting into the
SWBST discussed in the previous section. The PMA is able to accomplish this
using $O(1 + (\log{N}) /B)$ memory transfers. The proof is based on the fact
that each node has $\Theta(c^h)$ descendants (recall, this is the key property
that SWBST provided; shown in Equation~\ref{eq:bounds}). Thus, merging or splitting a node will cause
$\Theta(c^h)$ nodes to move, requiring $O(1 + c^h/B)$ memory transfers. It
follows from the definition of SWBST in \cite{swbst} that when a node is
rebalanced, it will require $\Theta(c^h)$ insertions or deletions until it is
rebalanced again. Thus, the amortized cost of each insertion and/or deletion is
$O(1/B)$. When a node is split, it may traverse all the way up the tree, and
thus the split/merge bound is $O((\log N) / B)$ memory transfers.

@@

In \cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
Oblivious String B-tree (COSB-tree), which works more efficiently
for variable length string keys. The COSB-tree uses a modified front-compression
scheme, which allows it to use less space and have improved
compressed data locality.

@@

The dynamic COSB-tree consists of three layers: a centroid tree, a PMA 
of hash data for the keys (known as the 
\emph{hashdata} layer), and a PMA of compressed values for the original key
 (known as the 
\emph{keydata} layer). Each leaf of the centroid tree contains a pointer into
the hashdata array. The centroid tree only contains a subset of the keys, so
each leaf points to the respective \emph{neighborhood} of the key being
searched for. After traversing the centroid tree and following the pointer into
the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
sequential scan is performed in the hashdata array layer to find $k$. Each
entry in the hashdata array contains the hash value of the corresponding key,
and a pointer to the compressed key value in the keydata layer. The keydata layer
is consulted to resolve hash collisions in the hashdata layer.

%a centroid tree at
%the top, a PMA structure of hash data for the keys known as the \emph{hashdata}
%layer; and on the bottom a PMA structure for the original key values,
%compressed using an augmented version of forward compression known as the
%\emph{keydata} layer. Each leaf of the centroid tree contains a pointer into
%the hashdata array. The centroid tree only contains a subset of the keys, so
%each leaf points to the respective \emph{neighborhood} of the key being
%searched for. After traversing the centroid tree and following the pointer into
%the hashdata array layer, we are within $O(\log{}N)$ keys of $k$. Then a
%sequential scan is performed in the hashdata array layer to find $k$. Each
%entry in the hashdata array contains the hash value of the corresponding key,
%and a pointer to the compressed key value in the keydata layer.

@@

\subsubsection{Centroid Trees} %todo:cite dictionary matching problem--sources in cosb-tree paper
\begin{figure}

@@

The centroid tree in the COSB-tree is layed out in memory using a modified vEB
(see Figure~\ref{fig:veb}) layout for weight-balanced binary trees. The
\emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into
equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x refers
to rounding x down to the nearest power of a power of 2. This ensures trees
of equal size on all \emph{levels of detail}, as the centroid tree is traversed.

********* LEILANI: Can you provide a 5 word description of levels of details ***

%The \emph{hyperhyperfloor} of x is
%denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to
%the nearest power of a power of 2. This ensures trees of equal size on all
%\emph{levels of detail}, as the centroid tree is traversed.
% Should explain what "level of detail" means here

@@

********* LEILANI: What is a centroid vertex? ***

@@

\subsubsection{Modifying Front Compression}

@@

\subsection{Operations}

@@

While B-trees achieve an optimal bound for the number of memory transfers for
search, they do not perform as well for bulk updates and insertions. The
buffered-repository tree (BRT) is an alternative to the B-tree that achieves an
improved amortized $O(\log{\frac{N}{B}})$ memory transfers for inserts compared
to the $O(\log_{B}{N})$ memory access of B-trees \cite{Buchsbaum}. In practice, log-structured
merge (LSM) trees \cite{ONeil96} are used for high-insert workloads.

@@

@@

\Search{} uses $O(\log_{B}{N})$ memory transfers where $N = \Theta(c^{F_k})$ and
$c$ is the fanout. This bound is achieved by searching over the
top and bottom recursive subtrees at each level when traversing the tree. A
key trick is that when you visit a node you only search its largest buffer
(the last buffer in the linked list for that level). The smaller buffers will
be visited eventually as you recurse down the tree. Eventually you will get
down to a single subtree that can fit into a single blocks and thus can be
searched with one memory transfer. Since the fanout is proportional to the
block size it will take us $O(\log_B N)$ block accesses until we get down to a
subtree and thus, by induction, the worst case runtime is $O(\log_B N)$.
